Hilbert space methods to approximate Gaussian processes using Stan

### 2019-02-26 Paper/manuscript/manuscript_05.tex
update GP section following Michael comments

A Gaussian process (GP) is a stochastic process which defines the distribution over a collection of random variables indexed by a continuous variable, i.e. $\left\lbrace f(t): t \in \mathcal{T}\right\rbrace$ for some index set $\mathcal{T}$. Gaussian processes have the defining property that the marginal distribution of every finite subset of random variables, $\left\lbrace f(t_1), f(t_2), \hdots, f(t_K) \right\rbrace$, is a multivariate Gaussian distribution.
\\
\\
In this work, Gaussian processes will take the role of prior distributions over function spaces for non-parametric regression in a Bayesian setting. 
Consider a data set $\mathcal{D} = \left\lbrace \mathbf{x}_n, y_n \right\rbrace_{n=1}^N$, where $y_n$ is a noisy observation of an unknown function $f$ at $\mathbf{x}_n \in \mathbb{R}^D$ , i.e. $y_n = f(\mathbf{x}_n) + \epsilon_n$. Our goal is to predict the value of the function $f^*=f(\mathbf{x}^*)$ evaluated at a new input point $\mathbf{x}^* \in \mathbb{R}^D$. That is, we want to obtain the predictive distribution $p(f^*|\mathcal{D})$ of $f^*$ conditioned on the data $\mathcal{D}$.
\\
\\
We assume a Gaussian process prior for $f \sim \mathcal{GP}(\mu(\mathbf{x}), k(\mathbf{x}, \mathbf{x}'))$, where $\mu: \mathbb{R} \rightarrow \mathbb{R}$ and $k: \mathbb{R} \times \mathbb{R} \rightarrow \mathbb{R}$ are the mean and covariance functions, respectively,
\begin{align}
 	\mu(\mathbf{x}) &= \mathbb{E}\!\left[f(\mathbf{x})\right],\\ 
 	k(\mathbf{x}, \mathbf{x}') &= \mathbb{E}\!\left[\left( f(\mathbf{x}) - \mu(\mathbf{x}) \right)\left( f(\mathbf{x}') - \mu(\mathbf{x}') \right)\right].
 \end{align} 
The mean and covariance functions completely characterize the distribution of $f$. 
\\
\\
{\color{red} $\bullet$ Introduce the joint model}
\begin{align}
	p(\mathbf{y}, \mathbf{f}, f^*) = p(\mathbf{y}|\mathbf{f})p(\mathbf{f}, f^*)
\end{align}
{\color{red} $\bullet$  argue that marginalizing and conditioning yields the predictive distribution of interest}
\begin{align}
	p(f^*|\mathbf{y}) = \frac{\int p(\mathbf{y}, \mathbf{f}, f^*)\text{d}\mathbf{f}}{p(\mathbf{y})}
\end{align}
{\color{red} $\bullet$ Give equations for predictive mean and variance for $f^*$ assuming Gaussian likelihood}
\\
\\
{\color{red} $\bullet$ Briefly standard computational challenges with GPs: cubic scaling and only closed form solution for Gaussian likelihoods. This should motivate the use of basis function expansions in a probabilistic programming setting such that with get both speed and modelling flexibility}
\\
\\

A GP model can be seen as a probability model over an infinite collection of random varibales. However, formally, the defining feature of a GP model is that the joint distribution of function values at any finite number of input values $\mathbf{x}$ is a multivariate normal distribution, 
%
\begin{equation}
f(\mathbf{x}) \sim \mathcal{N}(\mathbf{m},\mathbf{K})
\end{equation}

\noindent with $\mathbf{m}$ the mean vector and $\mathbf{K}$ the covariance matrix. The element $\mathbf{m}_i$ is given by the mean function $m(\mathbf{x}_i)$ and the element $\mathbf{K}_{ij}$ is given by the covariance function $k(x_i,x_j|\theta)$ which is a function of the inputs values $\mathbf{x}$ and the parameters $\theta$. The mean and covariance functions represent how the random functions behave on average and hoe the different points in the input space co-vary with respect to each other:

 In general the mean function $m(\mathbf{x})$ can be any function of the inputs, although this is usually set to zero for simplifying notation.

\vspace{0.5cm}
\noindent ... \\
... \\
...

\vspace{0.5cm}
A stationary covariance function is a function of $x-x'$, so we can write $k(x,x') = k(x-x')$. Isotropic covariance functions are those that are function of the distance between observations, $k(x,x') = k(|x-x'|) = k(r)$. The Mat\'ern family with half-integers is the class of isotropic covariance functions probably most commonly used, which are
%
\begin{eqnarray}
k_{\nu=\infty}(r)&=&\sigma^2 \text{exp}(-\frac{1}{2} r/\ell^2), \nonumber \\
k_{\nu=\frac{1}{2}}(r)&=&\sigma^2 exp(-r/\ell), \nonumber \\
k_{\nu=\frac{3}{2}}(r)&=&\sigma^2(1+\sqrt{3}r/\ell \text{exp}(-\sqrt{3}r/\ell), \nonumber \\
k_{\nu=\frac{5}{2}}(r)&=&\sigma^2(1+\sqrt{5}r/\ell+\frac{5}{3}r^2/\ell^2) \text{exp}(-\sqrt{5}r/\ell), \nonumber 
\end{eqnarray}

\noindent where $\nu$ is the order the kernel, and the $\ell$ and $\sigma$ are the lengt-scale and magnitud, respectively, of the kernel. The particular case where $\nu=\infty$ is commonly known as squared exponential (exponentiated quadratic) covariance function, and that with $\nu=1/2$ as exponential covariance function. These covariance functions can be easily generalized to the multidimensional case considering multidimensional inputs $\mathbf{r}$ and the multidimensional length-scale $\boldsymbol{\ell}$.

The spectral density functions associated with the univariate Mat\'ern covariance functions written above are
%
\begin{eqnarray}
S_{\nu=\infty}(w)&=& \sigma^2 \sqrt{2\pi} \cdot \ell \cdot \text{exp}(-0.5 \ell^2 w^2), \nonumber \\
S_{\nu=\frac{1}{2}}(w)&=& 2\sigma^2 \frac{1}{\ell}(\frac{1}{\ell^2} + \omega^2)^{-1}, \nonumber \\
S_{\nu=\frac{3}{2}}(w)&=& 4\sigma^2 \frac{\sqrt{3}}{\ell}^{3}(\frac{3}{\ell^2} + \omega^2)^{-2}, \nonumber \\
S_{\nu=\frac{5}{2}}(w)&=& \frac{16}{3}\sigma^2 \frac{\sqrt{3}}{\ell}^{5}(\frac{5}{\ell^2} + \omega^2)^{-3}, \nonumber 
\end{eqnarray}

\noindent where variable $w$ is in the frequency domain.

>>>>

A Gaussian process (GP) is a stochastic process which defines the distribution over a collection of random variables indexed by a continuous variable, i.e. $\left\lbrace f(t): t \in \mathcal{T}\right\rbrace$ for some index set $\mathcal{T}$. Gaussian processes have the defining property that the marginal distribution of every finite subset of random variables, $\left\lbrace f(t_1), f(t_2), \hdots, f(t_K) \right\rbrace$, is a multivariate Gaussian distribution.
\\
\\
In this work, Gaussian processes will take the role of prior distributions over function spaces for non-parametric regression in a Bayesian setting. 
Consider a data set $\mathcal{D} = \left\lbrace \mathbf{x}_n, y_n \right\rbrace_{n=1}^N$, where $y_n$ is a noisy observation of an unknown function $f$ at $\mathbf{x}_n \in \mathbb{R}^D$ , i.e. $y_n = f(\mathbf{x}_n) + \epsilon_n$. Our goal is to predict the value of the function $f^*=f(\mathbf{x}^*)$ evaluated at a new input point $\mathbf{x}^* \in \mathbb{R}^D$. That is, we want to obtain the predictive distribution $p(f^*|\mathcal{D})$ of $f^*$ conditioned on the data $\mathcal{D}$.
\\
\\
We assume a Gaussian process prior for the collection of function values $\mathbf{f}=\left\lbrace f(\mathbf{x}_n) \right\rbrace_{n=1}^N $, $\mathbf{f} \sim \mathcal{GP}(\mu(\mathbf{x}), k(\mathbf{x}, \mathbf{x}'))$, where $\mu: \mathbb{R}^D \rightarrow \mathbb{R}$ and $k: \mathbb{R}^D \times \mathbb{R}^D \rightarrow \mathbb{R}$ are the mean and covariance functions, respectively,
%
\begin{align*}
 	\mu(\mathbf{x}) &= \mathbb{E}\!\left[f(\mathbf{x})\right],\\ 
 	k(\mathbf{x}, \mathbf{x}') &= \mathbb{E}\!\left[\left( f(\mathbf{x}) - \mu(\mathbf{x}) \right)\left( f(\mathbf{x}') - \mu(\mathbf{x}') \right)\right].
\end{align*}

The mean and covariance functions represent how the random functions behave on average and how the different points in the input space co-vary with respect to each other. The mean and covariance functions completely characterize the multivariate Gaussian for $\mathbf{f}$, 
\begin{align*}
\mathbf{f} \sim \mathcal{N}(\boldsymbol{\mu},K),
\end{align*}

\noindent where $\boldsymbol{\mu} = \left\lbrace \mu(\mathbf{x}_n) \right\rbrace_{n=1}^N$ is the multivariate mean and $K$ the covariance matrix, where the element $K_{i,j}$ is given by the covariance function $k(\mathbf{x}_i,\mathbf{x}_j)$ at input values $\mathbf{x}_i$ and $\mathbf{x}_j$. The joint distribution of $\mathbf{f}$ and $f^*$ is also a multivariate Gaussian as,
%
\begin{eqnarray*}
p(\mathbf{f},\mathbf{f}')=\mathcal{N} \left( \left[ \begin{array}{cc}
\mathbf{f} \\ 
\mathbf{f}'
\end{array} \right] | 0,\left[ \begin{array}{cc}
K_{\mathbf{f},\mathbf{f}} & K_{\mathbf{f},f^*} \\ 
K_{f^*,\mathbf{f}} & K_{f^*,f^*}
\end{array} \right] \right),
\end{eqnarray*}

\noindent where $K_{\mathbf{f},f^*}$ defines the covariances between $\mathbf{f}$ and $f^*$, and $K_{f^*,f^*}$ defines the variance of $f^*$. By using the conditioning properties of a multivariate Gaussian, we can compute analytically the predictive distribution for $f^*$  given $\mathbf{f}$,
\begin{align*}
	p(f^*| \mathbf{f}) = \mathcal{N}(f^*| K_{f^*,\mathbf{f}} K_{\mathbf{f},\mathbf{f}}^{-1} \mathbf{f},  K_{f^*,f^*}-K_{f^*,\mathbf{f}} K_{\mathbf{f},\mathbf{f}}^{-1} K_{\mathbf{f},f^*}  )
\end{align*}

The joint distribution of observations $\mathbf{y} = \left\lbrace y_n \right\rbrace_{n=1}^N$ and function values $\mathbf{f}$ and $f^*$, $p(\mathbf{y}, \mathbf{f}, f^*)$, is the product of the conditional distribution for $\mathbf{y}$ given $\mathbf{f}$ and the joint distribution for $\mathbf{f}$ and $f^*$,
\begin{align*}
	p(\mathbf{y}, \mathbf{f}, f^*) = p(\mathbf{y}|\mathbf{f})p(\mathbf{f}, f^*)
\end{align*}

Integrating the joint distribution  $p(\mathbf{y}, \mathbf{f}, f^*)$ over the uncertainty of $\mathbf{f}$ and then conditioning on the observations $\mathbf{y}$ following the Bayes's rule yields the posterior predictive distribution for $f^*$ given the observations $\mathbf{y}$,
%
\begin{align*}
	p(f^*|\mathbf{y}) = \frac{\int p(\mathbf{y}|\mathbf{f})p(\mathbf{f}, f^*)\text{d}\mathbf{f}}{p(\mathbf{y})}
\end{align*}

If the observational model $p(\mathbf{y}|\mathbf{f})$ is Gaussian, the integrals over $\mathbf{f}$ can be computed analytically yielding,
\begin{align*}
	p(f^*| \mathbf{y}) = \mathcal{N}(f^*| K_{f^*,\mathbf{f}} (K_{\mathbf{f},\mathbf{f}} + \sigma^2I)^{-1} \mathbf{f},  K_{f^*,f^*}-K_{f^*,\mathbf{f}} (K_{\mathbf{f},\mathbf{f}} + \sigma^2I)^{-1} K_{\mathbf{f},f^*}  )
\end{align*}

In most modelling problems, however, the observational model is not Gaussian (e.g. binomial or multinomial classification, Poisson  regression, and many other modelling problems) and the posterior predictive distribution $p(f^*| \mathbf{y})$ is analytically untractable and can not be obtained in a closed-form expression. Approximate inference have to be used to tackle the intractable inference, tipically based on sampling methods such as MCMC \citep{brooks_2011} or analytical Gaussian approximations such as Laplace approximation (\citep{williams1998bayesian,rasmussen2006gaussian}, expectation propagation \citep{minka2001expectation}, or variational methods \citep{gibbs2000variational,csato2000efficient}. Furthermore, Gaussian processes in a direct implementation scales cubically in the number of sample points, $N$, which effectively limits the modelling inference for medium and large datasets. In the introduction section in this paper several methods tipically used for dealing with the computational complexity has been commented. 
\\
\\
In this paper we proposed an approximate model for Gaussian processes based on basis functions expansion which basically turns the GP prior model into a linear model. This linear GP model representation makes inference considerably faster and can be used as latent function in non-Gaussian observational models allowing modelling flexibility.
\\
\\
The covariance function is the crucial ingredient in a Gaussian process as it encodes our prior assumptions about the function, and defines a correlation structure which characterize the correlations between function values at different inputs. A stationary covariance function is a function of $\mathbf{x}-\mathbf{x}'$, so we can write $k(\mathbf{x},\mathbf{x}') = k(\mathbf{x}-\mathbf{x}')$, which means that the covariance is invariant to translations. Isotropic covariance functions are those that are function of the distance between observations, $k(\mathbf{x},\mathbf{x}') = k(||\mathbf{x}-\mathbf{x}'||) = k(\mathbf{r})$, which means that the covariacne is both translation and rotation invariant. The Mat\'ern family with half-integers is the class of isotropic covariance functions probably most commonly used, which are
%
\begin{eqnarray}
k_{\nu=\infty}(\mathbf{r})&=&\sigma^2 \text{exp}(-\frac{1}{2} \mathbf{r}/\ell^2), \nonumber \\
k_{\nu=\frac{1}{2}}(\mathbf{r})&=&\sigma^2 exp(-\mathbf{r}/\ell), \nonumber \\
k_{\nu=\frac{3}{2}}(\mathbf{r})&=&\sigma^2(1+\sqrt{3}\mathbf{r}/\ell \text{exp}(-\sqrt{3}\mathbf{r}/\ell), \nonumber \\
k_{\nu=\frac{5}{2}}(\mathbf{r})&=&\sigma^2(1+\sqrt{5}\mathbf{r}/\ell+\frac{5}{3}\mathbf{r}^2/\ell^2) \text{exp}(-\sqrt{5}\mathbf{r}/\ell), \nonumber 
\end{eqnarray}

\noindent where $\nu$ is the order the kernel, and the $\ell$ and $\sigma$ are the lengt-scale and magnitud, respectively, of the kernel. The particular case where $\nu=\infty$ is commonly known as squared exponential (exponentiated quadratic) covariance function, and that with $\nu=1/2$ as exponential covariance function. These covariance functions can be easily generalized to the use of multidimensional length-scale $\boldsymbol{\ell}$, which basically turns the isotropic covariance function into non-isotropic.

Stationary covariance functions can be represented in terms of their spectral densities \citep{Rasmussen_2006}. The spectral density functions associated with the Mat\'ern covariance functions written above are
%
\begin{eqnarray}
S_{\nu=\infty}(\mathbf{w})&=& \sigma^2 \sqrt{2\pi} \cdot \ell \cdot \text{exp}(-0.5 \ell^2 \mathbf{w}^2), \nonumber \\
S_{\nu=\frac{1}{2}}(\mathbf{w})&=& 2\sigma^2 \frac{1}{\ell}(\frac{1}{\ell^2} + \omega^2)^{-1}, \nonumber \\
S_{\nu=\frac{3}{2}}(\mathbf{w})&=& 4\sigma^2 \frac{\sqrt{3}}{\ell}^{3}(\frac{3}{\ell^2} + \omega^2)^{-2}, \nonumber \\
S_{\nu=\frac{5}{2}}(\mathbf{w})&=& \frac{16}{3}\sigma^2 \frac{\sqrt{3}}{\ell}^{5}(\frac{5}{\ell^2} + \omega^2)^{-3}, \nonumber 
\end{eqnarray}

\noindent where variable $\mathbf{w}$ is in the frequency domain.
